Gravity model and \textit{\textit{word2vec}}

### 2021-01-12 paper/current/notes/2021-0105-word2vec-vs-gravity-model/main.tex
edit some para

Let us approach the connection by initially focusing on the gravity model.
Suppose that flow of people between $N$ locations. The gravity model models the expected flow $\hat{T_{ij}}$ from location $i$ to $j$ by
\begin{align}
    \hat{T_{ij}} = C m_i m_j f(r_{ij}), \label{eq:gravity_model}
\end{align}
where $m_i$ is the population of locations of $i$, $r_{ij}$ is their distance, $f$ is a decreasing function of $r_{ij}$, and $C$ is a constant.
The choice of distance measure is not entirely unconstrained; it should have symmetricity (\ie $r_{ij} = r_{ji}$).
Apart from that, the distance measure $r_{ij}$ can be any dissimilarity measure even if it does not satisfy the axiom of metric space (such as triangle inequality $r_{ij}  + r_{jk} > r_{ik}$).
In the following, we refer to a measure of dissimilarity as distance for brevity.
In gravity model, the flow is symmetric, \ie $\hat{T_{ij}}=\hat{T_{ji}}$, which ensures the stationarity in population $m_{i}$.

Now, let us move our focus to the \textit{word2vec}.
In out setting, the model takes a sequence of locations and models the co-occurrence of center location $i$ and context location $j$ by
\begin{equation}
    p(j \given i) = \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{equation}
where $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector", respectively, and $Z_i=\sum_{x \in \mathcal{A}} \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant,
and $\mathcal{A}$ is the set of all locations.
Under restricting window size to be one, the expected flow between center location $a_i$ and context location $a_j$ is given by
\begin{align}
    \label{eq:flow_w2v}
    \hat{T_{ij}}=s_i P(j\given i) = \frac{s_i\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $s_i$ is the frequency of location $i$ summed for all trajectories in the dataset.
Although there are some similarities, Eq.~\ref{eq:flow_w2v} does not have a population term $m_i$. In additionm the distance can be asymmetric,$r_{ij}=- \bm{u}_{j} \cdot \bm{v}_{i} \neq - \bm{u}_{i} \cdot \bm{v}_{j}=r_{ji}$.

The gap between the gravity model and {\it word2vec} is filled by taking into account how we train the word2vec---\emph{the negative sampling}.
The negative sampling is an optimization method widely used for training the \textit{word2vec}, which operates as follows.
For each center--context word pair $(i,j)$, one generates $k$ noise center--context word pairs, denoted by $(i, \ell)$, where context word $\ell$ for the noise pair is sampled from a noise distribution, denoted by $p_{0}(\ell)$.
Given a center-context word pair, $(i,j)$, as an input, one predicts whether it comes from the original data or noise distribution based on a similarity between two words, \eg dot similarity in case of the \textit{word2vec}.

The negative sampling does not directly fit the word2vec to the data but through the synthetic classification task.
An often underappreciated fact is that the negative sampling has a bias and maximizes a different objective function~\autocite{Gutmann2010}.
With the negative sampling, the correct expression for the conditional probability is given by
\begin{align}
    P\left(j \given i \right):= \frac{p_0(j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}, \label{eq:prob_w2v_ng0}
\end{align}
where $Z'_i=\sum_{x \in \mathcal{A}} p_0(x) \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant.
See Section \ref{sec:correct_word2vec} for full calculations.

The \textit{word2vec} uses a noise distribution given by $p_{0}(j) = s^{\gamma} _j / \sum_{\ell'} s^{\gamma}_{\ell'}$, where $\gamma$ is a free parameter typically set to $3/4$ or $1$.
Parameter $\gamma=1$ is a special choice that ensures that the in-vector and out-vector are the same, \ie $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$, and substituting $\bm{u}_i = \bm{v}_j$ and $p_0(j) = s_j / \sum_{\ell'} s_{\ell'}$ into Eq~\eqref{eq:flow_w2v_ng0}, we have
\begin{align}
    P\left(j \given i \right) & := \frac{ p_0 (j) \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z'_i}                      \\
                              & =  \frac{s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z' _i\sum_{\ell'} s_{\ell'}},
\end{align}
which leads to the expected flow for the word2vec model, \ie
\begin{align}
    \hat{T_{ij}}=s_i P(j\given i) =  \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}}. \label{eq:flow_w2v_ng0}
\end{align}
In the word2vec, the flow $\hat{T_{ij}}_{ij}$ is symmetric by design (\ie $\hat{T_{ij}}=\hat{T_{ji}}$) because the word2vec neglects whether the context $j$ appears later or earlier from center $i$.
Therefore, we have
\begin{align}
    T_{ij} = T_{ji}
     & \Rightarrow
    \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}} =  \frac{s_j s _i  \exp(\bm{v}_i \cdot \bm{v}_{j})}{  Z'_j\sum_{\ell'} s_{\ell'}} \\
     & \Rightarrow
    Z'_i = Z'_j, \label{eq:euqal_rel}
\end{align}
suggesting that the normalization constant $Z' _i$ is constant.
Combining Eqs.~\eqref{eq:flow_w2v_ng0} and \eqref{eq:euqal_rel}, we have
\begin{align}
    \hat{T_{ij}}=  C s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i}),
\end{align}
where $C$ is a positive constant.
Eq.~\eqref{eq:flow_w2v_ng0} reveals the connection between the \textit{word2vec} and the gravity model, with with $d_{ij} = - \bm{u}_{j} \cdot \bm{v}_{i}$ and $f(x) = \exp(-x)$.
In sum, the \textit{word2vec} is a gravity model, in which the mass of location $i$ is its frequency in the trajectory data, and the distance between locations is measured by their dot similarities.

@@

The negative sampling does not directly fits a probability model to data but does so through the synthetic classification task.
This has a side effect: the negative sampling has a bias, and hence does not correctly train the model \cite{Dyer2014}.
The bias can be corrected using the noise contrastive estimation (NCE) \cite{Chia2010}, which solves the
same classification task but with a bias-corrected sigmoid function, \ie
\begin{align}
    \label{eq:sigmoid2}
    \frac{1}{1 + \exp\left( -\log f(z) + \log p_0(z) \right)}.
\end{align}
Note that Eq.~\eqref{eq:sigmoid2} has an additional term $\log p_0(z)$, which is the log-probability of $z$ for a noise distribution.

>>>>

Let us approach the connection by initially focusing on the gravity model.
Suppose that flow of people between $N$ locations. The gravity model models the expected flow $\hat{T_{ij}}$ from location $i$ to $j$ by
\begin{align}
    \hat{T_{ij}} = C m_i m_j f(r_{ij}), \label{eq:gravity_model}
\end{align}
where $m_i$ is the population of locations of $i$, $r_{ij}$ is their distance, $f$ is a decreasing function of $r_{ij}$, and $C$ is a constant.
In the following, we refer to a measure of dissimilarity as a distance for brevity.
In gravity model, the flow is symmetric, \ie $\hat{T_{ij}}=\hat{T_{ji}}$, which ensures the stationarity in population $m_{i}$.

Now, let us move our focus to the \textit{word2vec}.
In out setting, the model takes a sequence of locations and models the co-occurrence of center location $i$ and context location $j$ by
\begin{equation}
    p(j \given i) = \frac{\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{equation}
where $\bm{v}$ and $\bm{u}$ are the ``in-vector" and ``out-vector", respectively, and $Z_i=\sum_{x \in \mathcal{A}} \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant,
and $\mathcal{A}$ is the set of all locations.
Under restricting window size to be one, the expected flow between center location $a_i$ and context location $a_j$ is given by
\begin{align}
    \label{eq:flow_w2v}
    \hat{T_{ij}}=s_i P(j\given i) = \frac{s_i\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z_i},
\end{align}
where $s_i$ is the frequency of location $i$ summed for all trajectories in the dataset.
Although there are some similarities, Eq.~\ref{eq:flow_w2v} does not have a population term $m_i$. In additionm the distance can be asymmetric, $r_{ij}=- \bm{u}_{j} \cdot \bm{v}_{i} \neq - \bm{u}_{i} \cdot \bm{v}_{j}=r_{ji}$.

The gap between the gravity model and {\it word2vec} is filled by taking into account how we train the \textit{word2vec}---\emph{the negative sampling}.
The negative sampling is an optimization method widely used for training the machine learning model, which operates as follows.
For each center--context word pair $(i,j)$, one generates $k$ noise center--context word pairs, denoted by $(i, \ell)$, where context word $\ell$ for the noise pair is sampled from a noise distribution, denoted by $p_{0}(\ell)$.
Given a center-context word pair, $(i,j)$, as an input, one predicts whether it comes from the original data or noise distribution based on a similarity between two words, \eg dot similarity in case of the \textit{word2vec}.

The negative sampling does not directly fit the \textit{word2vec} to the data but through the synthetic classification task.
An often underappreciated fact is that negative sampling has a bias and maximizes a different objective function~\autocite{Gutmann2010}.
With the negative sampling, the correct expression for the conditional probability is given by
\begin{align}
    P\left(j \given i \right):= \frac{p_0(j)\exp(\bm{u}_j \cdot \bm{v}_{i})}{Z'_i}, \label{eq:prob_w2v_ng0}
\end{align}
where $Z'_i=\sum_{x \in \mathcal{A}} p_0(x) \exp(\bm{u}_{x} \cdot \bm{v}_{i})$ is a normalization constant, and a noise distribution given by $p_{0}(j) = s^{\gamma} _j / \sum_{\ell'} s^{\gamma}_{\ell'}$, where $\gamma$ is a free parameter typically set to $3/4$ or $1$.
See Section \ref{sec:correct_word2vec} for full calculations.

Parameter $\gamma=1$ is a special choice that ensures that the in-vector and out-vector are the same, \ie $\bm{u}_i = \bm{v}_i$ \autocite{levy2014neural}.
Setting $\gamma = 1$, and substituting $\bm{u}_i = \bm{v}_j$ and $p_0(j) = s_j / \sum_{\ell'} s_{\ell'}$ into Eq~\eqref{eq:flow_w2v_ng0}, we have
\begin{align}
    P\left(j \given i \right) & := \frac{ p_0 (j) \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z'_i}                      \\
                              & =  \frac{s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{ Z' _i\sum_{\ell'} s_{\ell'}},
\end{align}
which leads to the expected flow for the word2vec model, \ie
\begin{align}
    \hat{T_{ij}}=s_i P(j\given i) =  \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}}. \label{eq:flow_w2v_ng0}
\end{align}
In the \textit{word2vec}, the flow $\hat{T_{ij}}_{ij}$ is symmetric by design (\ie $\hat{T_{ij}}=\hat{T_{ji}}$) because the word2vec neglects whether the context $j$ appears later or earlier from center $i$.
Therefore, we have
\begin{align}
    T_{ij} = T_{ji}
     & \Rightarrow
    \frac{s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i})}{  Z'_i\sum_{\ell'} s_{\ell'}} =  \frac{s_j s _i  \exp(\bm{v}_i \cdot \bm{v}_{j})}{  Z'_j\sum_{\ell'} s_{\ell'}} \\
     & \Rightarrow
    Z'_i = Z'_j, \label{eq:euqal_rel}
\end{align}
suggesting that the normalization constant $Z' _i$ is constant.
Combining Eqs.~\eqref{eq:flow_w2v_ng0} and \eqref{eq:euqal_rel}, we have
\begin{align}
    \hat{T_{ij}}=  C s_i s _j  \exp(\bm{v}_j \cdot \bm{v}_{i}),
\end{align}
where $C$ is a positive constant.
Eq.~\eqref{eq:flow_w2v_ng0} reveals the connection between the \textit{word2vec} and the gravity model, with with $d_{ij} = - \bm{u}_{j} \cdot \bm{v}_{i}$ and $f(x) = \exp(-x)$.
In sum, the \textit{word2vec} is a gravity model, in which the mass of location $i$ is its frequency in the trajectory data, and the distance between locations is measured by their dot similarities.

@@

The negative sampling does not directly fit a probability model to data but does so through the synthetic classification task.
This has a side effect: the negative sampling has a bias and hence does not correctly train the model \cite{Dyer2014}.
The bias can be corrected using the noise contrastive estimation (NCE) \cite{Chia2010}, which solves the
same classification task but with a bias-corrected sigmoid function, \ie
\begin{align}
    \label{eq:sigmoid2}
    \frac{1}{1 + \exp\left( -\log f(z) + \log p_0(z) \right)}.
\end{align}
Note that Eq.~\eqref{eq:sigmoid2} has an additional term $\log p_0(z)$, which is the log-probability of $z$ for a noise distribution.
